How to find epistasis, Bitches!

### 2013-08-21 docs/protocols_ms/epistasis_protocols.tex
minor edits to first draft of a protocols manuscript

[Initially outline the challenges in the introduction. These can be address through the methods section] \\
\newline
1. Computational \\ 
\newline
[I think it would be worth re-discussing some of the computational problems of 2d scans. and highlighting the software + hardware solutions] \\
\newline
One of the first obstacles that make epistatic searches difﬁcult is the computational demands of the statistical analyses. When searching for independent additive effects, as is done for the majority of GWA studies, each SNP is tested for association with the phenotype. However, in order to most powerfully identify epistatic effects, the search must be increased to multiple dimensions (ref). Here the scale of the computational demand increases by $~(n^x)/2$ where $n$ is the number of SNPs and $x$ is the number of epistasis dimensions fitted. For example, testing for 2 loci interactions using SNP data from a 1 million SNP chip would require $1000000 * 999999/2 \approx 5e11$ individual tests. \\     
\newline 
2. Model choice \\
\newline
From a 2d 8df biallelic SNP model there are 4 epistatic variance components; additive x additive, additive x
dominance, dominance x additive and dominance x dominance. Within the literature there has been considerable discussion as to the likely degree of epistasis variance components and whether statistical power could be increased by parameterizing models for only a subset. We have shown then estimates of parameters representing the 4 variance components are proportionally represented amongst the significant epistatic pairs. This implies a full model is preferential, particularly if we are agonistic about the mechanism by which epistasis might arise. \\  
\newline
3. Statistical \\
\newline
A complete exhaustive scan of $m$ phenotypes and $n$ SNPs comprises of $((n*(n-1))^2/2)*m$ $8df$ F-tests. Given the high correlation structure inherent in genotype data as well as between multiple phenotypes, choices regarding multiple testing need to be carefully considered. [what can we say about tails of the distributions?]
[Perhaps this is a good point to outline out a 2-step procedure where main effects are initially corrected for] \\
The 8df F-test includes parameters for main (additive) effects of the two SNPs. Therefore it is important to test that test statistic isn't driven my marginal effects - something that is quite likely for a trait such as gene expression. Thus one should really consider a nested test.  \\
\newline
4. Interpretation / filtering \\
\newline
Things to consider and address during the analysis and output, particularly given the massive multiple testing and inability to easily evaluate the tails of the test-statistic distribution. Why we require at least 5 (or more) individuals in the smallest classt size for the genotypes. The choice of number is partly based total sample size. Assuming Hardy-Weinberg equilibrium, the frequency of the smallest class size is $p_1^2 * p_2^2$ where $p_1$ and $p_2$ are the frequencies of the minor alleles for SNPs 1 and 2 respectivly.      
\newline
\section{Methods}

@@

Outline the 2 stage models; 8df followed by (nested) 4df; or initial correction for main effects followed by a interection term. This would also be the place to point out the importance of a) running a full 2d scan rather than initially search for marginal effects and b) including all epistasis variance component parameters in the interection model.

>>>>

[Initially outline the challenges in the introduction. These can be address through the methods section] \\
\newline
1. Computational \\ 
\newline
[I think it would be worth re-discussing some of the computational problems of 2d scans. and highlighting the software + hardware solutions] \\
\newline
One of the first obstacles that make epistatic searches difﬁcult is the computational demands of the statistical analyses. When searching for independent additive effects, as is done for the majority of GWA studies, each SNP is tested for association with the phenotype. However, in order to most powerfully identify epistatic effects, the search must be increased to multiple dimensions (ref). Here the scale of the computational demand increases by $~(n^x)/2$ where $n$ is the number of SNPs and $x$ is the number of epistasis dimensions fitted. For example, testing for 2 loci interactions using SNP data from a 1 million SNP chip would require $1000000 * 999999/2 \approx 5e11$ individual tests. \\     
\newline 
2. Model choice \\
\newline
From a 2d 8df biallelic SNP model there are 4 epistatic variance components; additive x additive, additive x
dominance, dominance x additive and dominance x dominance. Within the literature there has been considerable discussion as to the likely degree of epistasis variance components and whether statistical power could be increased by parameterizing models for only a subset. We have shown then estimates of parameters representing the 4 variance components are proportionally represented amongst the significant epistatic pairs. This implies a full model is preferential, particularly if we are agonistic about the mechanism by which epistasis might arise. \\  
\newline
3. Statistical \\
\newline
A complete exhaustive scan of $m$ phenotypes and $n$ SNPs comprises of $((n*(n-1))^2/2)*m$ $8df$ F-tests. Given the high correlation structure inherent in genotype data as well as between multiple phenotypes, choices regarding multiple testing need to be carefully considered. [what can we say about tails of the distributions?]
[Perhaps this is a good point to outline out a 2-step procedure where main effects are initially corrected for] \\
The 8df F-test includes parameters for main (additive) effects of the two SNPs. Therefore it is important to test that test statistic isn't driven my marginal effects - something that is quite likely for a trait such as gene expression. Thus one should really consider a nested test.  \\
\newline
4. Interpretation / filtering \\
\newline
Things to consider and address during the analysis and output, particularly given the massive multiple testing and inability to easily evaluate the tails of the test-statistic distribution. Why we require at least 5 (or more) individuals in the smallest class size for the genotypes. The choice of number is partly based total sample size. Assuming Hardy-Weinberg equilibrium, the frequency of the smallest class size is $p_1^2 * p_2^2$ where $p_1$ and $p_2$ are the frequencies of the minor alleles for SNPs 1 and 2 respectively.      
\newline
\section{Methods}

@@

Outline the 2 stage models; 8df followed by (nested) 4df; or initial correction for main effects followed by a interaction term. This would also be the place to point out the importance of a) running a full 2d scan rather than initially search for marginal effects and b) including all epistasis variance component parameters in the interaction model.
